Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-14 paper_jasa.tex
working on picking pvalue assumption section

Lemma \ref{lemma} helps us estimate the plot signal strength based on which the visual test decision could be made based on multiple evaluations on a lineup. From the equation \ref{fig:pval_plot_signal} following from Lemma \ref{lemma} we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives an estimate as  $$\hat p_D= (1- u/K)/(m-1)$$ where $K$ is the total number of independent evaluations on a lineup with $u$ being number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with classical test $p$-values. Notice that plot signal strength is equivalent to classical test $p$-value when $p$-value is less than or equal to 0.05 and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.

Figure \ref{fig:pval_pcorrect} shows evidence of direct association of percentage of correct evaluations with classical test $p$-values. When $p$-value is larger than 0.15, it is almost impossible to correctly identify the observed plot in the lineup. We observed this for both experiment 1 and 2. For experiment 3, the scenario is different as in this case people might have observed something other than the effect $E$.

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_percent_correct.pdf}}
       \caption{Percentage of correct responses decreases rapidly with the increase of p-value. After p-value exceeds 0.15 it is very unlikely to identify the actual plot. The theoretical justification of this is shown in figure \ref{fig:pval_power}. \blue{The scenario is different for experiment 3. $p$-value does not seem to have any effect on percent correct. This may be due to Type-III error.} }
       \label{fig:pval_pcorrect}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Classical test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$) estimated using Lemma \ref{lemma}. The solid line is a 45 degree line which represents the equity shown in equation \ref{eqn:power_estimate}.}
       %\hh{Mahbub, could you exchange this plot for one with  $p_D$ estimated as $(1 - \#  correct/ \# shown)/(m-1)$ as given in lemma 3.1 now?}}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizntally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the exeriment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.) Red indicates the actual data plot, which is not always the plot in the lineup with the lowest $p$-value. As $\beta$ gets larger (stronger effect) people tended to select the plot with the lowest $p$-value.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_rank_counts12.pdf}}
       \caption{Rank of p-values in a lineup of size 20 and the number of subjects choosing that plot in the lineup of experiment 1 and 2 with raw data (screening criteria 1). This indicates that most of the subjects pick the plot with minimum p-value (rank 1). There are subjects choosing the lineups with highest p-value(rank 20) but we can attribute this behavior to the type-I error. Can we?  \green{Maybe don't need this plot}}
       \label{fig:P-val_rank}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_log_counts.pdf}}
       \caption{Frequency of plot picks by log$_{10}$ $p$-value in all of the sixty (60) individual lineups for experiment 1. The colored lines indicate the $p$-value of actual data plot.}
       \label{fig:P-val_log}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.5}{\includegraphics{p_val_log_counts2.pdf}}
       \caption{Frequency of plot picks by log$_{10}$ $p$-value in all of the seventy (70) individual lineups for experiment 2. The colored lines indicate the $p$-value of actual data plot.}
       \label{fig:P-val_log2}
\end{figure*}

>>>>

\green{third experiment doesn't really have a p-value, does it? Need to explain this. Why are there still so many small p-values in experiment 3?}

In the real setting, where visual inference is to be useful, there will be no classical tests or traditional $p$-values. Yet, assessing the strength of the structure is a critical component of inference. For experiments 1 and 2, to assess the signal strength we examine the proportion of correct picks, with multiple users and the $p$-value of the actual data plot. For experiment 3 the \green{How is p-value calculated here?} Figure \ref{fig:pval_pcorrect} displays scatterplots of these two variables for each experiment. As the $p$-value increases the proportion of correct responses falls, evidence of direct association between proportion of correct responses and classical test $p$-values. When $p$-value is larger than 0.15, it is almost impossible to correctly identify the observed plot in the lineup. We observed this for both experiment 1 and 2. For experiment 3, the scenario is different, because there are two structures, the association between variables, and the cluster of contaminated points.% as in this case people might have observed something other than the effect $E$.

\green{Should we use proportion of correct, rather than percentage? Wouldn't that work into the equations better?}

Yet, it is possible to obtain an estimate the strength of the signal in the actual data plot, when there are multiple evaluations, using Lemma \ref{lemma}. From Equation \ref{fig:pval_plot_signal}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
\[
\hat p_D= \frac{1- \frac{u}{K}}{m-1}
\]
where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with classical test $p$-values. Notice that plot signal strength is equivalent to classical test $p$-value when $p$-value is less than or equal to 0.05 and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_percent_correct.pdf}}
       \caption{Percentage of correct responses decreases rapidly with the increase of p-value. After p-value exceeds 0.15 it is very unlikely to identify the actual plot. The theoretical justification of this is shown in figure \ref{fig:pval_power}. \blue{The scenario is different for experiment 3. $p$-value does not seem to have any effect on percent correct. This may be due to Type-III error.} \green{p-value on log scale?} }
       \label{fig:pval_pcorrect}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Classical test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$) estimated using Lemma \ref{lemma}. The solid line is a 45 degree line which represents the equity shown in equation \ref{eqn:power_estimate}. \green{p-value on log scale?}}
       %\hh{Mahbub, could you exchange this plot for one with  $p_D$ estimated as $(1 - \#  correct/ \# shown)/(m-1)$ as given in lemma 3.1 now?}}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizntally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the exeriment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.) Red indicates the actual data plot, which is not always the plot in the lineup with the lowest $p$-value. In both experiments as $\beta$ gets larger (stronger effect) people tended to select the plot with the lowest $p$-value. This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. The results are clearer for experiment 2, that used a continuous covariate. So for most observers, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

Experiment 3 does not have $p$-values associated with each plot, so the assumption can only be tested with the first two experiments. Practically, this is not an assumption that is explicitly made, because there will not be a $p$-value available. Rather it is assumed that the observer has the ability to select the plot with the most extreme structure. A few of the lineups indicated that occasionally people cue on different patterns, for example, in experiment 1 the lineup $\beta=0, n=100, sd=5, rep=1$ (first row, fourth column) many subjects chose the same plot, that had a higher $p$-value. Some of these anomalous lineups were selcted for follow-up studies using an eyetracker to investigate what people were looking at.

@@

\begin{figure*}[hbtp]
   \centering
       \includegraphics[width=0.95\textwidth]{p_val_rank_counts12.pdf}
       \caption{Rank of p-values in a lineup of size 20 and the number of subjects choosing that plot in the lineup of experiment 1 and 2 with raw data (screening criteria 1). This indicates that most of the subjects pick the plot with minimum p-value (rank 1). There are subjects choosing the lineups with highest p-value(rank 20) but we can attribute this behavior to the type-I error. Can we?  \green{Maybe don't need this plot}}
       \label{fig:P-val_rank}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \includegraphics[width=0.95\textwidth]{p_val_log_counts.pdf}
       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the sixty (60) individual lineups for experiment 1. The colored lines indicate the plot with the lowest $p$-value. Rows correspond to $\beta$, smallest to largest values from top to bottom. Columns correspond to other experimental treatments, sample size $(n)$, error standard deviation $(\sigma)$, and replicate. Empty cells indicate no lineup for that combination of parameters. Highest counts tend to be the plot in the lineup having the lowest $p$-value.}
       \label{fig:P-val_log}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \includegraphics[width=0.95\textwidth]{p_val_log_counts2.pdf}
       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the seventy (70) individual lineups for experiment 2. The colored lines indicate the plot with the lowest $p$-value. Rows correspond to $\beta$, smallest to largest values from top to bottom. Columns correspond to other experimental treatments, sample size $(n)$, error standard deviation $(\sigma)$, and replicate. Empty cells indicate no lineup for that combination of parameters. Highest counts tend to be the plot in the lineup having the lowest $p$-value, even more so than for experiment 1.}
       \label{fig:P-val_log2}
\end{figure*}
